---
title: "Summarizing with dplyr"
author: "Amy L"
date: "6/5/2020"
output: html_document
---

##Section 3 Overview  
- be able to use the summarize() and group_by() verb in **dplyr** to facilitate summarzing data  
- be able to access values by using the dot placeholder  
- be able to use arrange() to examine data after sorting  

### Dplyr  
**Key points**    
- summarize() computes summary statistics from the data frame and returns a data frame whose **column** names are defined within the function call  
- summarize() can compute any summary function that operates on vectors and returns a single value, but it CANNOT operate on functions that return multiple values  
- summarize() is aware of variable names within data frames and can use them directly 
```{r}
library(tidyverse) #note that dpylr is part of tidyverse pacakage
library(dslabs)
data(heights)

#Compute avg and st. dev for males
s <- heights %>% filter(sex == "Male") %>% summarize(average = mean(height), standard_deviation = sd(height))

#access average and standard deviation from summary table
s$average
s$standard_deviation

#compute median, min , and max
heights %>% filter(sex == "Male") %>% summarize(median = median(height), minimum = min(height), maximum = max(height))

#alternative way to get min, median, max in base R
quantile(heights$height, c(0, 0.5, 1))

#generates an error: summarize can only take functions that return a SINGLE value
##heights %>% filter(sex == "Male") %>% summarize(range = quantile(height, c(0, 0.5, 1)))
```


### The Dot Placeholder
**Key Points**  
* The dot operator allows you to access values stored in data that is being piped in using %>% character  
* The dot is a placeholder for the data being passed in through the pipe  
* The dot operator allows dplyr functions to return single vectors or numbers instead of only data frames  
* us_murder_rate %>% .$rate is equivalent to us_murder_rate$rate  
* Note that an equivalent way to extract a single column using the pipe is us_murder_rate %>% pull(rate)
```{r}
library(tidyverse)
library(dslabs)
data(murders)

murders <- murders %>% mutate(murder_rate = total/population *100000)

summarize(murders, mean(murder_rate))

###METHOD 1###
#calculate US murder rate, generating a data frame
us_murder_rate <- murders %>% summarize(rate = sum(total)/sum(population) * 100000)
us_murder_rate #3.03

#extract the numeric US murder rate with the dot operator
us_murder_rate %>% .$rate  #3.03

###METHOD 2###
#calculate and extract the murder rate with one pipe
us_murder_rate <- murders %>% summarize(rate = sum(total)/sum(population) * 100000) %>% .$rate
us_murder_rate
```

### Group By  
**Key Points**  
* The group_by() function from dplyr converts a data frame to a grouped data frame, creating groups using one or more variables.  
* Summarize() and some other dplyr functions will behave *differently* on grouped data frames  
* Using summarize() on a grouped data frame computes the summary statistics for EACH of the separate groups 
```{r}
#load the libaries 
library(tidyverse) #loads dplyr
library(dslabs)
data(heights)
data(murders)

#compute separate average and st.deviation for male/female heights
heights %>% group_by(sex) %>% summarize(average = mean(height), standard_deviation = sd(height))

#compute median murder rate in 4 regions of country 
murders <- murders %>% mutate(murder_rate = total/population * 100000)

murders %>% group_by(region) %>% summarize(median_rate = median(murder_rate))

```

### Sorting Data Tables  
- The arrange() function from dplyr sorts a data frame by a given column 
- Be default, arrange() sorts in ascending order(lowest to highest). To instead sort in descending order, use the function desc() inside of arrange()  
- You can arrange() by multiple levels: within equivalent values of the first level, observations are sorted by the second level, and so on  
- The top_n() function shows the top results ranked by a given variable, but the results are not ordered. BUT, you can combine top_n() with arrange() to return the top results in order.
```{r}
#libraries and data
library(tidyverse)
library(dslabs)
data(murders)

#set up murders object (i.e., add new column of "murder rates")
murders <- murders %>% mutate(murder_rate = total/population*100000) 

#arrange by population column, smallest to largest
murders %>% arrange(population) %>% head()

#arrange by murder rate, smalleset to largest
murders %>% arrange(murder_rate) %>% head()

#arrange by murder rate in descending order
murders %>% arrange(desc(murder_rate)) %>% head()

#arrange by region alphabetically, then by murder rate within each region
murders %>% arrange(region, murder_rate) %>% head()

#show the top 10 states with highest murder rate, not ordered by rate
murders %>% top_n(10, murder_rate)

#show the top 10 states with highest murder rate, ordered by rate
murders %>% arrange(desc(murder_rate)) %>% top_n(3)
```

###Datacamp Exercises
```{r}
library(NHANES)
data(NHANES)
library(dslabs)
data(na_example)

#note: summarization function in R will return NA if ANY of the entries of the input vector is an NA
mean(na_example) 
sd(na_example)

#To ignore the NAs, use the na.rm argument:
mean(na_example, na.rm = TRUE)
sd(na_example, na.rm = TRUE)
```

####Blood pressure exercise
```{r}
library(dplyr)
library(NHANES)
data("NHANES")

#Q1) - Filter the NHANES dataset so that only 20-29 year old females are included and assign this new data frame to the object tab.
tab <- NHANES %>% filter(AgeDecade == " 20-29" & Gender == "female") %>% head(); tab

#Q2) - Now we will compute the average and standard deviation for the subgroup we defined in the previous exercise (20-29 year old females),
ref <- NHANES %>% filter(AgeDecade == " 20-29" & Gender == "female") %>% summarize(average = mean(BPSysAve, na.rm = TRUE), standard_deviation = sd(BPSysAve, na.rm = TRUE)); ref

#Q3) - Generate only the average blood pressure for 20-29 year old females using place holder (.)
## modify the code we wrote for previous exercise.

ref_avg <- NHANES %>% filter(AgeDecade == " 20-29" & Gender == "female") %>% summarize(average = mean(BPSysAve, na.rm = TRUE),standard_deviation = sd(BPSysAve, na.rm=TRUE)) %>% .$average; ref_avg

#<------OR----->
#ref_avg <- NHANES %>% filter(AgeDecade == " 20-29" & Gender == "female") %>% summarize(average = mean(BPSysAve, na.rm = TRUE),standard_deviation = sd(BPSysAve, na.rm=TRUE)) %>% pull(average)

#Q4) - Calculate two other data summaries: the minimum and the maximum for the BPSysAve variable
NHANES %>% filter(AgeDecade == " 20-29"  & Gender == "female") %>% summarize(min=min(BPSysAve,na.rm=TRUE),max=max(BPSysAve,na.rm=TRUE))

#Q5) - Compute the average and st.deviation of systolic blood pressure for females for EACH age group separately 
NHANES %>% filter(Gender == "female") %>% group_by(AgeDecade) %>% summarize(average=mean(BPSysAve,na.rm=TRUE),standard_deviation = sd(BPSysAve,na.rm=TRUE))

#Q6) - Calculate the average and st. deviation of systolic blood pressure for males 
NHANES %>% filter(Gender == "male") %>% group_by(AgeDecade) %>% summarize(average=mean(BPSysAve,na.rm=TRUE),standard_deviation = sd(BPSysAve,na.rm=TRUE))

#Q7) - We can actually combine both of these summaries into a single line of code because group_by permits us to group by more than one variable -> group_by(AgeDecade, Gender) is used to group by both age decades and gender 
NHANES %>% group_by(AgeDecade, Gender) %>% summarize(average=mean(BPSysAve,na.rm=TRUE),standard_deviation = sd(BPSysAve,na.rm=TRUE))

#8) - Use Arrange() to explore differences in systolic blood pressure across races
NHANES %>% arrange(BPSysAve) #arranges ascending systolic bp
NHANES %>% arrange(desc(BPSysAve)) #arranges descending sys. bp

#Compare the systolic blood pressure across values of the Race1 variable for males between the ages of 40-49. Compute the avg and st.dev for each value of Race1 for males in the age decade 40-49. Order the result table from lowest to highest average sysolic BP.

NHANES %>% filter(AgeDecade ==" 40-49" & Gender == "male") %>% group_by(Race1) %>% summarize(average=mean(BPSysAve,na.rm=TRUE),standard_deviation = sd(BPSysAve,na.rm=TRUE))%>% arrange(average)
```
[Here is a website with notes and answers!](https://rpubs.com/faisalcep/dsVisualization)
