---
title: "Gapminder"
author: "Amy L"
date: "6/6/2020"
output: html_document
---

##Section 4 Overview  
- Understand how Hans Rosling and the Gapminder Foundation use effective data visualization to convey data-based trends  
- Be able to apply the **ggplot2** techniques from previous section to answer questions using data  
- Understand how fixed scales across plots can ease comparisons 
- Be able to modify graphs to improve data visualzaiton 


###Case Study: Trends in World Health and Economics  
* Data visualization can be used to dispel common myths and educate the public and contradict sensationalist or outdated claims and stories  
* Use real world data to answer questions about world health and economics  
- 1)Is it still fair to consider the world as divided into the West and the developing world?  
- 2)Has income inequality across countries *worsened* over the last 40 years?  

###Gapminder Dataset  
```{r}
# load and inspect gapminder data
library(tidyverse)
library(dslabs)
data(gapminder)
head(gapminder)

# compare infant mortality in Sri Lanka and Turkey
gapminder %>%
    filter(year == 2015 & country %in% c("Sri Lanka", "Turkey")) %>%
    select(country, infant_mortality)
```

###Life Expectancy and Fertility Rates  
- A Scatterplot of life expectancy versus fertility in 1962 suggests that the view "the world  is divided into the western world and the developing world" was grounded in reality 50 years ago. Is it still the case today?  

```{r}
ds_theme_set()
filter(gapminder, year == 1962) %>% ggplot(aes(fertility, life_expectancy)) + geom_point()

#Use different color for the continents
filter(gapminder, year == 1962) %>% ggplot(aes(fertility, life_expectancy, color = continent)) + geom_point()

#NOTE: that if we just change the year, the scale of the plot changed tooï¼ --> Thus... we need to use facet!
filter(gapminder, year == 2012) %>% ggplot(aes(fertility, life_expectancy, color = continent)) + geom_point()
```

###Faceting  
- Faceting makes multiple side-by-side plots stratified by some variable -> this is a way to ease comparisons  
- The **facet_grid()** function allows faceting by UP to two variables, with rows faceted by one variable and columns faceted by the other variable. To facet by only ONE variable, use the dot operator as the other variable.
- The **facet_wrap()** function facets by one variable and automatically wraps the series of plots so they have readable dimensions.  
- Faceting KEEPS the axes fixed across all plots, easing comparisons between plots.  
- **Answer**: The data suggest that the developing versus Western world view no longer makes sense in 2012!

```{r}
# facet by continent AND year
filter(gapminder, year %in% c(1962, 2012)) %>% ggplot(aes(fertility, life_expectancy, col = continent)) +  geom_point() + facet_grid(continent ~ year)

# facet by year only
filter(gapminder, year %in% c(1962, 2012)) %>% ggplot(aes(fertility, life_expectancy, col = continent)) + geom_point() + facet_grid(. ~ year)

# facet by year, plots wrapped onto multiple rows
years <- c(1962, 1980, 1990, 2000, 2012)
continents <- c("Europe", "Asia")
gapminder %>% filter(year %in% years & continent %in% continents) %>% ggplot(aes(fertility, life_expectancy, col = continent)) + geom_point() + facet_wrap(~year)

```

###Time Series Plots  
- Time series plots have time on the x-axis and a variable of interest on the y-axis.  
- The geom_line() geometry connects adjacent data point to form a **continuous** line. A line plot is appropriate when points are regularly spaced, densely packed and form a single data series.  
- You can plot multiple lines on the same graph. Remember to group or color by a variable so that the lines are plotted independently.  
- Labeling is usually preferred over legends.* However, legends are easier to make and appear by default. Add a label with geom_text(), specifying the coordinates where the label should appear on the graph.  
```{r}
#Code: single time series
# scatterplot of US fertility by year
gapminder %>%
    filter(country == "United States") %>%
    ggplot(aes(year, fertility)) +
    geom_point()

# line plot of US fertility by year
gapminder %>%
    filter(country == "United States") %>%
    ggplot(aes(year, fertility)) +
    geom_line()


#Code: Multiple time series 
# line plot fertility time series for two countries- only one line (Incorrect!!!)
countries <- c("South Korea", "Germany")
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, fertility)) +
    geom_line()
    
# line plot fertility time series for two countries - one line per country (correct!!!)
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, fertility, group = country)) +
    geom_line()

# fertility time series for two countries - lines colored by country
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, fertility, col = country)) +
    geom_line() #col = color


#Code: Adding text labels to a plot
# life expectancy time series - lines colored by country and labeled
labels <- data.frame(country = countries, x = c(1975, 1965), y = c(60, 72))
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, life_expectancy, col = country)) +
    geom_line() + geom_text(data = labels, aes(x, y, label = country), size = 5) + theme(legend.position = "none")
```


###Transformations  
**KEY POINTS**  
- We use GDP data to compute income in US dollars, adjusted for inflation.    
- Log transformations convert multiplicative changes into additive changes.  
- Common transformations are the log base 2 transformation and the log base 10 transformation. The choice of base depends on the range of the data. The natural log is NOT recommended for visualization because it is difficult to interpret.  
- The mode of a distribution is the value with the highest frequency. The mode of a **normal** distribution is the average. A distribution can have multiple *local* modes.  
- There are two ways to use log transformations in plots: 1) transform the data before plotting or 2) transform the axes of the plot. 
> Log scales have the advantage of showing the original values as axis labels, while log transformed values ease interpretation of intermediate values between labels.  

- Scale the x-axis using scale_x_continuous() or scale_x_log10() layers in **ggplot2**.  
- In 1970, income distribution is bimodal, consistent with the dichotomous Western vs. developing worldview.  
- Notes: Gross Domestic Product(GDP) is the monetary value of all finished goods and services made within a country during a specific period. GDP per capita = GDP/population

```{r}
#add dollars per DAY variable 
gapminder <- gapminder %>% mutate(dollars_per_day = gdp/population/365)

#histogram of dollars per day
past_year <- 1970

gapminder %>% filter(year == past_year & !is.na(gdp)) %>% ggplot(aes(dollars_per_day)) + geom_histogram(binwidth = 1, color = "black")

#repeat histogram with log2 scaled DATA
gapminder %>% filter(year == past_year & !is.na(gdp)) %>% ggplot(aes(log2(dollars_per_day))) + geom_histogram(binwidth = 1, color = "black")

#repeat histogram with log2 scaled X-Axis
gapminder %>% filter(year == past_year & !is.na(gdp)) %>% ggplot(aes(dollars_per_day)) + geom_histogram(binwidth = 1, color = "black") + scale_x_continuous(trans = "log2")
```

###Stratify and Boxplot  
**Key Points**  
- Make boxplots stratified by a categorical variable using the geom_boxplot() geometry.  
- Rotate axis labels by changing the theme through element_text(). You can change the angle and justification of the text labels.  
- Consider ordering your factors by a meaningful value with reorder(), which changes the order of factor levels based on a **related** numeric vector. This is a way to ease comparisons.  
- Show data by adding data points to the boxplot with a geom_point() layer. This adds information beyond the five-number summary to your boxplot. However, too many data points can obfuscate your message. 

**CODE: Boxplot of GDP by region**  
```{r}
# add dollars per day variable
gapminder <- gapminder %>%
    mutate(dollars_per_day = gdp/population/365)
    
# number of regions
length(levels(gapminder$region))

# boxplot of GDP by region in 1970
past_year <- 1970
p <- gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    ggplot(aes(region, dollars_per_day))
p + geom_boxplot()

# rotate names on x-axis
p + geom_boxplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

**CODE: Simple example of the reorder function**  
```{r}
# by default, factor order is alphabetical
fac <- factor(c("Asia", "Asia", "West", "West", "West"))
levels(fac)

# reorder factor by the category means
value <- c(10, 11, 12, 6, 4)
fac <- reorder(fac, value, FUN = mean)
levels(fac)
```


**Code: Enhanced boxplot ordered by median income, scaled, and showing data points**
```{r}
# reorder by median income and color by continent
p <- gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    mutate(region = reorder(region, dollars_per_day, FUN = median)) %>%  # reorder
    ggplot(aes(region, dollars_per_day, fill = continent)) +    # color by continent
    geom_boxplot() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + xlab("")
p

# log2 scale y-axis
p + scale_y_continuous(trans = "log2")

# add data points
p + scale_y_continuous(trans = "log2") + geom_point(show.legend = FALSE)
```



###Comparing Distributions

###Density Plots

###Ecological Fallacy 
