---
title: 'Data Science: Visualization'
author: "Amy L"
date: "5/27/2020"
output:
  html_document: default
  pdf_document: default
---

### 1.2 Introduction to Distributions 
```{r}
# load the dataset
library(dslabs)
data(heights)
# make a table of category proportions
prop.table(table(heights$sex))
```
a <- seq(min(my_data), max(my_data), length = 100)    # define range of values spanning the dataset
cdf_function <- function(x) {    # computes prob. for a single value
  mean(my_data <= x)
}
cdf_values <- sapply(a, cdf_function)
plot(a, cdf_values)

#### Normal Distribution
```{r}
# define x as vector of male heights
library(tidyverse)
library(dslabs)
data(heights)
index <- heights$sex=="Male"
x <- heights$height[index]

# calculate the mean and standard deviation manually
average <- sum(x)/length(x)
SD <- sqrt(sum((x - average)^2)/length(x))

# built-in mean and sd functions - note that the audio and printed values disagree
average <- mean(x)
SD <- sd(x) 
#NOTE: The built-in R function sd() calculates the st. dev, but it divides by length(x)-1 instead of length(x). This is neglible if the sample is large. 
c(average = average, SD = SD)

# calculate standard units - converts a vector of approximately normally distributed values into z-scores.
z <- scale(x)

# calculate proportion of values within 2 SD of mean
mean(abs(z) < 2)
```


#### Using the function pnorm()
```{r}
#Given male heights x, we can estimate the probability that a male is TALLER than 70.5 inches with:
library(tidyverse)
library(dslabs)
data(heights)
x <- heights %>% filter(sex=="Male") %>% pull(height)
1 - pnorm(70.5, mean(x), sd(x))

# plot distribution of exact heights in data
#table(x)
#prop.table(table(x))
plot(prop.table(table(x)), xlab = "a = Height in inches", ylab = "Pr(x = a)")

# probabilities in actual data over length 1 ranges containing an integer with normal approxim.
mean(x <= 68.5) - mean(x <= 67.5)
pnorm(68.5, mean(x), sd(x)) - pnorm(67.5, mean(x), sd(x))

mean(x <= 69.5) - mean(x <= 68.5)
pnorm(69.5, mean(x), sd(x)) - pnorm(68.5, mean(x), sd(x))

# probabilities in actual data over other ranges DON't match normal approx as well
mean(x <= 70.9) - mean(x <= 70.1)
pnorm(70.9, mean(x), sd(x)) - pnorm(70.1, mean(x), sd(x))

#Estimate the proportion of seven footers in the NBA; assume height with mean of 69 and std of 3
#in the whole population and that 1 billion people are between ages 18:40 (age in order to be in NBA)
p <- 1 - pnorm(7*12, 69, 3)
N <- round(p * 10^9)
10/N #about 3.5% of 7-footers the age of 18:40 are in the NBA
```


### 1.3 Quantiles, Percentiles, and Boxplots
Load the heights dataset from the dslabs package:
```{r}
library(dslabs)
data(heights)
```

Use summary() on the heights$height variable to find the quartiles:
```{r}
summary(heights$height) 
```

Find the percentiles of heights$height: (think of standardilized tests such as the SAT)
```{r}
p <- seq(0.01, 0.99, 0.01);p
percentiles <- quantile(heights$height, p); percentiles #percentile is a named numeric vector 
#Note: given a dataset data and desired quantile q, you can find the qth quantile of data with: quantile(data,q)
#The qth quantile is the value at which  ð‘ž % of the observations are equal to or less than that value.
#NOTE: The 50th percentile is also known as the median.
```

Confirm that the 25th and 75th percentiles match the 1st and 3rd quartiles. Note that quantile() returns a named vector. You can access the 25th and 75th percentiles like this (adapt the code for other percentile values):
```{r}
percentiles[names(percentiles) == "25%"] #66  "25% of the observations are equal to or less than 66""
percentiles[names(percentiles) == "75%"]
percentiles
```

#### Finding quantiles with qnorm()
The qnorm() function gives the theoretical value of a quantile with probability p of observing a value equal to or less than that quantile value given a normal distribution with mean mu and standard deviation sigma:
```{r}
qnorm(0.975) #Answer = 1.96 --> 97.5% of values are less than 1.96
```

The pnorm() function gives the probability that a value from a standard normal distribution will be less than or equal to a z-score value z.
```{r}
pnorm(1.96)  #probability that a value is less than or equal to 1.96 is 0.975
```
Note that qnorm() and pnorm() are inverse functions
```{r}
pnorm(qnorm(0.975))
```

You can use qnorm() to determine the theoretical quantiles of a dataset: that is, the theoretical value of quantiles assuming that a dataset follows a normal distribution. Run the qnorm() function with the desired probabilities p, mean mu and standard deviation sigma. 
```{r}
p <- seq(0.01, 0.99, 0.01)
theoretical_quantiles <- qnorm(p, 69, 3); theoretical_quantiles
```
*Note: Theoretical quantiles can be compared to sample quantiles determined with the quantile function in order to evaluate whether the sample follows a normal distribution.

#### [Quantile-Quantile Plots](https://rafalab.github.io/dsbook/distributions.html#quantile-quantile-plots)
Quantile-quantile plots, or QQ-plots, are used to check whether distributions are well-approximated by a normal distribution.
Given a proportion p, the quantile q is the value such that the proportion of values in the data below q is p.
In a QQ-plot, the sample quantiles in the observed data are compared to the theoretical quantiles expected from the normal distribution. 

**If the data are well-approximated by the normal distribution, then the points on the QQ-plot will fall near the identity line (sample = theoretical).**
- Calculate sample quantiles (observed quantiles) using the quantile() function.
- Calculate theoretical quantiles with the qnorm() function. qnorm() will calculate quantiles for the standard normal distribution (Î¼=0,Ïƒ=1) by default, but it can calculate quantiles for any normal distribution given mean() and sd() arguments.

```{r}
# define x and z
library(tidyverse)
library(dslabs)
data(heights)
index <- heights$sex=="Male" #index is a logical vector 
x <- heights$height[index]; x
z <- scale(x) #scale() returns the Z-score

# proportion of data below 69.5
mean(x <= 69.5)

# calculate observed and theoretical quantiles
p <- seq(0.05, 0.95, 0.05)
observed_quantiles <- quantile(x, p)
theoretical_quantiles <- qnorm(p, mean = mean(x), sd = sd(x)) #theoretical_quantitles assumes normal distribution

# make QQ-plot
plot(theoretical_quantiles, observed_quantiles)
abline(0,1) #abline() adds a straight line through the current plot (intercept:0, slope =1)

# make QQ-plot with scaled values
observed_quantiles <- quantile(z, p)
theoretical_quantiles <- qnorm(p)
plot(theoretical_quantiles, observed_quantiles)
abline(0,1) #y=x


#DATACAMP - Percentiles
library(dslabs)
data(heights)
quantile(heights$height, seq(0.01, 0.99, 0.01))

male <- heights$height[heights$sex=="Male"]
female <- heights$height[heights$sex=="Female"]
male_percentiles <- quantile(male, seq(0.1, 0.9, 0.2)); male_percentiles
female_percentiles <- quantile(female, seq(0.1, 0.9, 0.2));female_percentiles
#dataframe
df <- data.frame(female = female_percentiles, male = male_percentiles);df
```

### 1.4 Exploratory Data Analysis
- If a distribution is not normal, it CANNOT be summarized with only the mean and standard deviation --> use a histogram, smooth density or boxplot instead. 
> "The greatest value of a picture is when it forces us to notice what we never expected to see."- Johh Tukey 